---
title: "NF1 Thesis Analysis"
author: "Marshall McQuillen"
date: "12/10/2018"
output: 
  pdf_document: 
    latex_engine: xelatex
    toc: true
    toc_depth: 4
---

```{r, include=FALSE, eval=TRUE}
# set global options
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE)
```

# Study Aims

1. To assess whether parental knowledge of NF1 is dependent on the child's mode of inheritance of the condition (familial vs de novo mutations)

2. To determine how exposure to genetic counseling affects knowledge of NF1.

```{r test_setup}
# load packages
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(ltm))

# load & setup data
grand.data <- read.csv("/Users/marshallmcquillen/Data_Science/NF1_Thesis/data/grand_data.csv")
grand.data$mode_of_inheritance <- ifelse(grand.data$does_your_childs_other_parent_have_nf1. == "No" & grand.data$do_you_have_nf1. == "No", "De Novo", "Inherited")

# original significance level
alpha <- 0.05

# Bonferroni Correction
adj.alpha <- alpha / 4

# create subsets on whether respondent has seend a genetic counselor
grand.data$seen_gc <- grand.data$have_you_ever_met_with_a_genetic_counselor.
seen.gc <- subset(grand.data,
                  seen_gc == "Yes")
not.seen.gc <- subset(grand.data,
                      seen_gc == "No")

# create subsets based on whether the respondent has NF1
group.a <- subset(grand.data, group_id == "Group A")
group.b <- subset(grand.data, group_id == "Group B")
group.c <- subset(grand.data, group_id == "Group C")
```

\pagebreak

# Data Overview & Cleaning

There were a total of 303 responses to the questionare:

* 11 respondents were "Not sure" if they had NF1. These respondents were dropped from the data set since both study aims are contingent upon knowing this information.

* One respondent skipped the question, "Does your child's other parent have NF1?" This respondent was kept since they answered the question, "Do you have NF1?" and their answer to "Does your child's other parent have NF1?" was imputed as "No."

* One respondent skipped the question, "Doe you have NF1?" This respondent was dropped from the data set since both study aims are contingent upon knowing this information.

* 14 respondents do not have NF1 and are "Not sure," if the childs other parent has NF1. These respondents were dropped from the data set since both aims of the thesis are dependent on this information.

* Three additional respondents were dropped from the study due to leaving other questions that will be discussed blank.

```{r group_summary_setup}
summary.df <- data.frame(Mean = c(mean(group.a$score),
                                  mean(group.b$score),
                                  mean(group.c$score)),
                         StandardDeviation = c(sd(group.a$score),
                                               sd(group.b$score),
                                               sd(group.c$score)))
rownames(summary.df) <- c('Group A','Group B','Group C')


# removing the "first" two ages from the age range (10 & 15), since I suspect these respondents thought the question
# was regarding the age of the diagnosed children
age.range <- range(sort(grand.data$age)[3:length(grand.data$age)], na.rm = TRUE)
age.mean <- mean(grand.data$age, na.rm = TRUE)
age.sd <- sd(grand.data$age, na.rm = TRUE)
```

After the above adjustments, the final data set contains 274 respondents, with an age ranging from `r age.range[1]` to `r age.range[2]` and a mean age of `r age.mean`. The standard deviation of the ages is `r age.sd`.

# Testing Methodology

Although there are only two separate study aims, the first will require a total of 3 pairwise statistical tests, comparing the average test scores of the three groups below to one another using Welch's $t$-Test.

* **Group A** - Respondents who do not have NF1 and the child's other parent also does not have NF1.

* **Group B** - Respondents who do not have NF1 and the child's other parent does have NF1.

* **Group C** - Respondents who do have NF1 and the child's other parent does not have NF1.

These three tests, combined with the test for the second study aim, create a total of 4 statistical tests. In order to control the family wise error rate, the original significance level ($\alpha = 0.05$) was adjusted using the Bonferroni Correction, reducing the level of statistical significance to $0.05 / 4 = 0.0125$.

A table with summar statistcs of the three groups is displayed below.

```{r group_summary_table}
knitr::kable(summary.df,
             caption = "Test Score Statistics")
```

\pagebreak

# Study Aim 1 - Does the mode of inheritance (familial vs de novo mutations) of the child have an effect on parental knowledge of NF1?

## Test 1

* $H_0:$ Respondents in Group B do not have a different average test score than respondents in Group A.

* $H_a:$ Respondents in Group B have a different average test score than respondents in Group A.

\hfill

```{r}
# welch's t-test
test.1.results <- t.test(group.a$score, group.b$score, var.equal = FALSE, alternative = 'two.sided')

test.1.data <- rbind.data.frame(group.a, group.b)

subtitle <- paste("Mean's of each group displayed as vertical lines: Group A =", round(mean(group.a$score), 4), "|", "Group B =", round(mean(group.b$score), 4))

#tiff(filename = '../images/GroupAvsGroupB.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#    res = 300)
ggplot(test.1.data, aes(score, fill = group_id)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(group.a$score), color = 'darkgreen') +
    geom_vline(xintercept = mean(group.b$score), color = 'orange') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    scale_fill_manual(values = c('darkgreen','orange')) +
    ggtitle("Group A vs. Group B",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")
#dev.off()

print(paste("Test 1 p value =", test.1.results$p.value))
```

Welch's two-tailed $t$-Test returns the above p value, with which we fail to reject the null hypothesis.

See the file GroupAvsGroupB.png, in the images directory.

\hfill

## Test 2

* $H_0:$ Respondents in Group A do not have a higher average test score than respondents in Group C.

* $H_a:$ Respondents in Group A have a higher average test score than respondents in Group C.

\hfill

```{r}
# welch's t-test
test.2.results <- t.test(group.a$score, group.c$score, var.equal = FALSE, alternative = 'greater')

test.2.data <- rbind.data.frame(group.a, group.c)

subtitle <- paste("Mean's of each group displayed as vertical lines: Group A =", round(mean(group.a$score), 4), "|", "Group C =", round(mean(group.c$score), 4))

#tiff(filename = '../images/GroupAvsGroupC.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(test.2.data, aes(score, fill = group_id)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(group.a$score), color = 'darkgreen') +
    geom_vline(xintercept = mean(group.c$score), color = 'purple') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    scale_fill_manual(values = c('darkgreen','purple')) +
    ggtitle("Group A vs. Group C",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")
#dev.off()

print(paste("Test 2 p value =", test.2.results$p.value))
```

Welch's one-sided $t$-Test returns the above p value. Being well below the Bonferroni corrected significance level of `r adj.alpha`, the null hypothesis is rejected in favor of the alternative.

See the file GroupAvsGroupC.png, in the images directory.

\hfill

## Test 3

* $H_0:$ Respondents in Group B do not have a higher average test score than respondents in Group C.

* $H_a:$ Respondents in Group B have a higher average test score than respondents in Group C.

\hfill

```{r}
# welch's t-test
test.3.results <- t.test(group.b$score, group.c$score, var.equal = FALSE, alternative = 'greater')

test.3.data <- rbind.data.frame(group.b, group.c)

subtitle <- paste("Mean's of each group displayed as vertical lines: Group B =", round(mean(group.b$score), 4), "|", "Group C =", round(mean(group.c$score), 4))

#tiff(filename = '../images/GroupBvsGroupC.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(test.3.data, aes(score, fill = group_id)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(group.b$score), color = 'orange') +
    geom_vline(xintercept = mean(group.c$score), color = 'purple') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    scale_fill_manual(values = c('orange','purple')) +
    ggtitle("Group B vs. Group C",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")
#dev.off()

print(paste("Test 3 p value =", test.3.results$p.value))
```

Welch's one-sided $t$-Test returns the above p value. Being well below the Bonferroni corrected significance level of `r adj.alpha`, the null hypothesis is rejected in favor of the alternative.

See the file GroupBvsGroupC.png, in the images directory.

\pagebreak

# Study Aim 2 - Does exposure to genetic counseling affect knowledge of NF1?

## Test 4

* $H_0:$ Respondents who have seen a genetic counselor do not have a higher average test score than those that have not seen a genetic counselor.

* $H_a:$ Respondents who have seen a genetic counselor have a higher average test score than those that have not seen a genetic counselor

\hfill

```{r}
# welch's t-test
test.4.results <- t.test(seen.gc$score, not.seen.gc$score, var.equal = FALSE,
       alternative = 'greater')

test.4.data <- rbind.data.frame(seen.gc, not.seen.gc)

subtitle <- paste("Mean's of each group displayed as vertical lines: Yes =", round(mean(seen.gc$score), 4), "|", "No =", round(mean(not.seen.gc$score), 4))

#tiff(filename = '../images/SeenGCvsNotSeenGC.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(test.4.data, aes(score, fill = seen_gc)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(not.seen.gc$score), color = '#F8766D') +
    geom_vline(xintercept = mean(seen.gc$score), color = '#00BFC4') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    ggtitle("Have you seen a Genetic Counselor?",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")
#dev.off()

print(paste("Test 4 p value =", test.3.results$p.value))
```

Welch's one-sided $t$-Test returns the above p value. Being well below the Bonferroni corrected significance level of `r adj.alpha`, the null hypothesis is rejected in favor of the alternative.

See the file SeenGCvsNotSeenGC.png, in the images directory.

```{r cronbachs_alpha, include=FALSE}
quiz.questions <- c('one_half_or_50._of_genetic_information_is_passed_down_from_mother_to_child',
                    'there_is_more_than_one_gene_that_causes_nf1',
                    'a_father_can_pass_down_an_nf1_gene_mutation_to_his_daughters',
                    'all_people_who_have_an_nf1_gene_mutation_will_develop_cancer',
                    'nf1_symptoms_vary_from_one_person_to_another',
                    'a_person_is_born_with_nf1',
                    'nf1_can_skip_generations',
                    'if_a_woman_with_nf1_has_one_child_with_nf1_her_second_child_will_definitely_not_have_nf1',
                    'people_with_nf1_are_generally_more_likely_to_develop_cancer_at_a_younger_age',
                    'if_a_woman_with_nf1_has_scoliosis_then_her_child_will_also_develop_scoliosis',
                    'tumors_in_nf1_can_appear_anywhere_in_the_body',
                    'people_with_nf1_from_different_families_will_always_have_different_symptoms',
                    'cafe.au.lait_spots_.brown_marks_on_the_skin._are_often_the_first_sign_that_a_person_has_nf1',
                    'a_baby_with_nf1_may_be_born_with_a_tumor',
                    'all_cases_of_nf1_can_be_detected_in_a_womans_pregnancy_by_ultrasound',
                    'half_of_people_with_nf1_have_a_family_history_of_nf1',
                    'a_person_with_nf1_can_develop_tumors_that_may_lead_to_vision_loss_or_blindness',
                    'women_with_nf1_are_at_an_increased_risk_for_breast_cancer',
                    'is_scoliosis_associated_with_nf1.',
                    'is_attention_deficit_hyperactivity_disorder_associated_with_nf1.',
                    'are_clubbed_feet_associated_with_nf1.',
                    'are_congenital_heart_defects_associated_with_nf1.',
                    'are_seizures_associated_with_nf1.',
                    'are_bumps_on_the_skin_associated_with_nf1.',
                    'are_allergies_associated_with_nf1.',
                    'is_high_blood_pressure_associated_with_nf1.',
                    'are_learning_disabilities_associated_with_nf1.',
                    'are_optic_gliomas_associated_with_nf1.',
                    'is_infertility_associated_with_nf1.',
                    'are_lisch_nodules_.dark_spots_on_the_iris._the_colored_part_of_the_eye._associated_with_nf1.',
                    'is_small_head_size_associated_with_nf1.',
                    'are_cataracts_associated_with_nf1.',
                    'how_is_nf1_diagnosed.',
                    'what_types_of_exams_may_be_helpful_when_trying_to_make_a_diagnosis_of_nf1.',
                    'what_causes_nf1.',
                    'what_is_the_function_of_the_nf1_gene_in_the_body.',
                    'are_men_or_women_more_likely_to_be_affected_by_nf1.',
                    'how_many_copies_of_a_non.working_nf1_gene_must_one_have_to_be_affected_with_nf1.',
                    'if_a_person_with_nf1_has_a_child_what_is_the_chance_that_the_child_will_have_nf1.',
                    'what_is_the_increased_risk_for_a_person_with_nf1_to_develop_cancer_in_their_lifetime_compared_to_someone_without_nf1.',
                    'during_childhood_what_kind_of_screening_may_be_done_for_optic_gliomas_.eye_tumors..',
                    'how_do_the_symptoms_of_nf1_change_as_a_person_gets_older.',
                    'what_is_the_most_concerning_symptom_of_a_tumor_to_suggest_it_has_become_malignant_.cancerous..')

quiz.df <- grand.data[, quiz.questions]
cronbachs <- cronbach.alpha(quiz.df)$alpha
```

# Cronbach's Alpha

Cronbach's alpha is a statistic used to measure the reliability of a set of test items, or more specifically, how well a set of test items (questions) measures the underlying concept which the test is attempting to assess. In the context, of this thesis, the Cronbach's Alpha can be interpreted as a numeric summary of how well the knowledge based questions assess a respondents true knowledge level of NF1.

Using 43 quiz questions returned a Cronbach's Alpha value of `r cronbachs`.

\pagebreak

# Test Limitations

As with any statistical test, there are limitations imposed by the methods I have chosen to use. In contrast to Student's $t$-test, Welch's $t$-test does not assume that the variances of the two samples are equal. However, the assumption that the data is Normally distributed is maintained.

The Shapiro-Wilk test is a statistical test used to determine whether that particular sample came from a population that is normally distributed. The null hypothesis of this test is that the sample came from a Normally distributed population, the alternative therefore being that the sample did not originate from a Normally distributed population. As shown in the table below, all samples have a "low" p value from the Shapiro-Wilk test, groups B and C being the only ones that are above the Bonferroni adjusted significance threshold.

```{r}
group.a.normality <- shapiro.test(group.a$score)$p.value
group.b.normality <- shapiro.test(group.b$score)$p.value
group.c.normality <- shapiro.test(group.c$score)$p.value
seen.gc.normality <- shapiro.test(seen.gc$score)$p.value
not.seen.gc.normality <- shapiro.test(not.seen.gc$score)$p.value

summary.df <- data.frame(PValue = c(group.a.normality,
                                    group.b.normality,
                                    group.c.normality,
                                    seen.gc.normality,
                                    not.seen.gc.normality))

rownames(summary.df) <- c('Group A','Group B','Group C','Seen Genetic Counselor','Not Seen Genetic Counselor')
knitr::kable(summary.df,
             caption = 'Shapiro-Wilk Test P-Values')
```

This, however, leads to another important topic, the problem of multiple comparisons. When one decides to correct for multiple comparisons, as I have done here using the Bonferroni Correction, the goal is to reduce the probability of a false positive. This practice *necessarily* increases the probability of a false negative. I adjusted the original significance level of 0.05 by accounting for the 4 statistical tests that addressed the study aims of this thesis. However, had I included the above 5 statistical tests that might fall under the heading of *Assumption Testing*, the adjusted significance level would fall to `r alpha/9`. Clearly, deciding where the bounds of a "family" of statistical tests begin and end has a large impact on the whether a particular test is deemed "significant," and, "there is no firm rule on this."$^{[1]}$

The problem of multiple comparisons is an area of active research and, seeing as, "There is no universally accepted approach for dealing with the problem of multiple comparisons,"$^{[1]}$ deciding on the Bonferroni Correction was one of many possible solutions to the problem.

\pagebreak

# Demographic Analysis

```{r proportion_of_respondents_who_have_nf1}
#tiff(filename = '../images/HaveNF1.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(grand.data, aes(do_you_have_nf1., fill = do_you_have_nf1.)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Do you have have NF1?") +
    ylab("Proportion")
#dev.off()


result <- table(grand.data$do_you_have_nf1.)
knitr::kable(result,
             caption = "Do you have NF1?",
             col.names = c('Response','Count'))
```

The plot associated with the above table is HaveNF1.png, in the images directory.

\hfill

```{r proportion_of_respondents_who_have_met_a_gc}
#tiff(filename = '../images/MetGC.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(grand.data, aes(have_you_ever_met_with_a_genetic_counselor.,
                       fill = have_you_ever_met_with_a_genetic_counselor.)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Have you ever met with a Genetic Counselor?") +
    ylab("Proportion")
#dev.off()

result <- table(grand.data$have_you_ever_met_with_a_genetic_counselor.)
knitr::kable(result,
             caption = "Have you ever met with a Genetic Counselor?",
             col.names = c('Response','Count'))
```

The plot associated with the above table is MetGC.png, in the images directory.

\hfill

```{r employment_status}
grand.data$employment_status <- sub("^$", "Unknown", grand.data$employment_status)

display.df <- as.data.frame(grand.data %>%
    count(employment_status) %>%
    arrange(desc(n))) %>%
    mutate(employment_status = factor(employment_status, employment_status))

#tiff(filename = '../images/EmploymentStatusDistribution.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(display.df, aes(x = reorder(employment_status, -n),
                       y = n,
                       fill = employment_status)) +
    geom_bar(aes(y = n/sum(display.df$n)), stat = 'identity') +
    geom_text(aes(y = n/sum(display.df$n),
                  label = scales::percent(n/sum(display.df$n))),
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Employment Status") +
    ylab("Proportion")
#dev.off()

result <- table(grand.data$employment_status)
knitr::kable(result,
             caption = "What is your employment status?",
             col.names = c('Response','Count'))
```

The plot associated with the above table is EmploymentStatusDistribution.png, in the images directory.

\pagebreak

```{r gender}
grand.data$gender <- sub("^$", "Unknown", grand.data$gender)

#tiff(filename = '../images/GenderDistribution.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(grand.data, aes(gender,
                       fill = gender)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Gender") +
    ylab("Proportion")
#dev.off()

result <- table(grand.data$gender)
knitr::kable(result,
             caption = "Gender",
             col.names = c('Response','Count'))
```

The plot associated with the above table is GenderDistribution.png, in the images directory.

\hfill

```{r ethnicity}
grand.data$ethnicity <- sub("^$", "Unkown", grand.data$ethnicity)

display.df <- as.data.frame(grand.data %>%
    count(ethnicity) %>%
    arrange(desc(n))) %>%
    mutate(ethnicity = factor(ethnicity, ethnicity))

#tiff(filename = '../images/EthnicityDistribution.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(display.df, aes(x = reorder(ethnicity, -n),
                       y = n,
                       fill = ethnicity)) +
    geom_bar(aes(y = n/sum(display.df$n)), stat = 'identity') +
    geom_text(aes(y = n/sum(display.df$n),
                  label = scales::percent(n/sum(display.df$n))),
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Ethnicity") +
    ylab("Proportion")
#dev.off()


result <- table(grand.data$ethnicity)
knitr::kable(result,
             caption = "Ethnicity",
             col.names = c('Response \t','Count'))
```

The plot associated with the above table is EthnicityDistribution.png, in the images directory.

\pagebreak

One person did not answer the question regarding their educational background, and therefore was excluded from the below plot.

\hfill

```{r educational_background}
levels(grand.data$educational_background) <- c('Unkown',
                                               levels(grand.data$educational_background)[2:length(levels(grand.data$educational_background))])
grand.data[is.na(grand.data$educational_background), "educational_background"] <- 'Unkown'

display.df <- as.data.frame(grand.data %>%
                                group_by(mode_of_inheritance) %>%
                                count(educational_background))

inherited <- display.df[display.df$mode_of_inheritance != 'De Novo', ] %>%
    mutate(percentage = n/sum(n)) %>%
    arrange(desc(percentage)) %>%
    mutate(educational_background = factor(educational_background, educational_background))

de.novo <- display.df[display.df$mode_of_inheritance == 'De Novo', ] %>%
    mutate(percentage = n/sum(n)) %>%
    arrange(desc(percentage)) %>%
    mutate(educational_background = factor(educational_background, educational_background))
                            
display.df <- rbind(de.novo, inherited)

#tiff(filename = '../images/EducationDistribution.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(display.df, aes(x = educational_background,
                       y = percentage,
                       fill = educational_background)) +
    geom_bar(stat = 'identity') +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = percentage,
                  label = scales::percent(percentage)),
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Educational Background") +
    ylab("Proportion")
#dev.off()

result <- table(grand.data$educational_background)
knitr::kable(result[2:length(result)],
             caption = "Educational Background",
             col.names = c('Response \t','Count'))
```

The plot associated with the above table is EducationDistribution.png, in the images directory.

\hfill

It is important to note that the answers to the below question, "What specialists does your affected child see?" are not mutually exclusive - children who see a Neurologist may have seen an Oncologist and a Geneticist as well. 

\hfill

```{r specialists_seen}
cols = c("does_your_oldest_child_see_a_neurologist_for_nf_care.",
         "does_your_oldest_child_see_an_oncologist_for_nf_care.",
         "does_your_oldest_child_see_a_geneticist_for_nf_care.",
         "does_your_oldest_child_see_an_ent.audiologist_for_nf_care.",
         "does_your_oldest_child_see_a_pediatrician_for_nf_care.",
         "does_your_oldest_child_see_a_dermatologist_for_nf_care.",
         "mode_of_inheritance")

display.df <- grand.data %>%
    group_by(mode_of_inheritance) %>%
    summarise(Neurologist = sum(does_your_oldest_child_see_a_neurologist_for_nf_care.),
              Oncologist = sum(does_your_oldest_child_see_an_oncologist_for_nf_care.),
              Geneticist = sum(does_your_oldest_child_see_a_geneticist_for_nf_care.),
              Audiologist = sum(does_your_oldest_child_see_an_ent.audiologist_for_nf_care.),
              Pediatrician = sum(does_your_oldest_child_see_a_pediatrician_for_nf_care.),
              Dermatologist = sum(does_your_oldest_child_see_a_pediatrician_for_nf_care.))
df.d <- melt(display.df, id.vars='mode_of_inheritance')

de.novo <- df.d[df.d$mode_of_inheritance == 'De Novo',] %>%
    arrange(desc(value)) %>%
    mutate(variable = factor(variable, variable))

inherited <- df.d[df.d$mode_of_inheritance != 'De Novo',] %>%
    arrange(desc(value)) %>%
    mutate(variable = factor(variable, variable))

df.d <- rbind(de.novo, inherited)

#tiff(filename = '../images/SpecialistsByModeOfInheritance.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(df.d, aes(x = reorder(variable, -value),
                 y = value,
                 fill = variable)) +
    geom_bar(stat = 'identity') +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = value, label = value),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("What specialists does your affected child see?",
            "Faceted by mode of inheritance") +
    ylab("Count")
#dev.off()
```

See the file SpecialistsByModeOfInheritance.png, in the images directory.

\hfill

```{r frequency_of_visits}
#tiff(filename = '../images/FrequencyOfVisitsByModeOfInerheritance.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(grand.data, aes(how_often_does_your_.oldest._child_see_an_nf_doctor.,
                       y = (..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))) +
    geom_bar(aes(fill = how_often_does_your_.oldest._child_see_an_nf_doctor.)) +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(label = paste(round(((..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))*100,2),'%', sep = '')),
              stat = "count",
              vjust = -0.25,
              size = 3) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("How often does your oldest child see a NF1 doctor?") +
    ylab("Proportion")
#dev.off()

result <- table(grand.data$how_often_does_your_.oldest._child_see_an_nf_doctor., grand.data$mode_of_inheritance)
knitr::kable(result,
             caption = "How often does your oldest child see a NF1 doctor?")
```

The plot associated with the above table is FrequencyOfVisitsByModeOfInerheritance.png, in the images directory.

\pagebreak

```{r frequency_of_visits_vs_score}
display.df <- grand.data %>%
    group_by(how_often_does_your_.oldest._child_see_an_nf_doctor.) %>%
    summarise(mean_score = mean(score))
df.d <- melt(display.df, id.vars='how_often_does_your_.oldest._child_see_an_nf_doctor.')

#tiff(filename = '../images/FrequencyOfVisitsAgainstQuizScore.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(df.d, aes(x = how_often_does_your_.oldest._child_see_an_nf_doctor., y = value, fill = how_often_does_your_.oldest._child_see_an_nf_doctor.)) +
    geom_bar(stat = 'identity') +
    scale_y_continuous(labels = scales::percent) +
    geom_text(aes(y = value, label = paste(round(value*100, 2), '%', sep = '')),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("How often does your oldest child see a NF1 doctor?") +
    ylab("Mean Quiz Score")
#dev.off()
```

See the file FrequencyOfVisitsAgainstQuizScore.png, in the images directory.

\hfill

```{r mode_of_inheritance_vs_seen_gc}
#tiff(filename = '../images/MetGCByModeOfInheritance.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(grand.data, aes(have_you_ever_met_with_a_genetic_counselor.,
                       fill = have_you_ever_met_with_a_genetic_counselor.)) +
    geom_bar(aes(y = (..count..))) +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = (..count..),
                  label = (..count..)),
              stat = "count",
              vjust = -0.25) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Have you ever met with a Genetic Counselor?",
            "Faceted by mode of inheritance") +
    ylab("Count")
#dev.off()
```

See the file MetGCByModeOfInheritance.png, in the images directory.

\hfill

```{r severity_symptoms}
#tiff(filename = '../images/SeverityOfSymptomsByModeOfInheritance.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(grand.data,
       aes(how_would_you_describe_the_severity_of_symptoms_of_your_.oldest._child_with_nf1.,
           y = (..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))) +
    geom_bar(aes(fill = how_would_you_describe_the_severity_of_symptoms_of_your_.oldest._child_with_nf1.)) +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(label = paste(round(((..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))*100,2),'%', sep = '')),
              stat = "count",
              vjust = -0.25,
              size = 3) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("How would you describe the severity of symptoms\nof your oldest child with NF1?",
            "Faceted by mode of inheritance") +
    ylab("Proportion")
#dev.off()

result <- table(grand.data$how_would_you_describe_the_severity_of_symptoms_of_your_.oldest._child_with_nf1., grand.data$mode_of_inheritance)
knitr::kable(result,
             caption = "How would you describe the severity of symptoms of your oldest child with NF1?")
```

The plot associated with the above table is SeverityOfSymptomsByModeOfInheritance.png, in the images directory.

\hfill

Similar to one of the previous plots, it is important to note that the answers to the below question, "Where do you obtain knowledge regarding NF1 should you have questions?" are not mutually exclusive - parents who ask their doctor might also reference online sources.

\hfill

```{r obtain_knowledge}
display.df <- grand.data %>%
  group_by(mode_of_inheritance) %>%
  summarize(
    Doctor = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_your_doctor.),
    NF1Organization = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_an_nf_organization_website.),
    FamilyMembers = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_family_members_with_nf1.),
    OnlineSources = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_online_searches.),
    SocialMedia = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_social_media_sites_.such_as_facebook..),
    OtherFamilies = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_other_families_you_know_that_have_nf1.),
    DontInvestigate = sum(if_you_have_questions_regarding_nf1_do_you_not_obtain_additional_information.))
display.df <- melt(display.df, 'mode_of_inheritance')

levels(display.df$variable) <- c('Doctor', 'NF1 Organization', 'Family Members', 'Online Sources', 'Social Media', 'Other Families', "Dont Investigate")

de.novo <- display.df[display.df$mode_of_inheritance == 'De Novo',] %>%
    arrange(desc(value)) %>%
    mutate(variable = factor(variable, variable))

inherited <- display.df[display.df$mode_of_inheritance != 'De Novo',] %>%
    arrange(desc(value)) %>%
    mutate(variable = factor(variable, variable))

display.df <- rbind(de.novo, inherited)

#tiff(filename = '../images/ObtainKnowledgeByModeOfInheritance.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(display.df, aes(x = variable,
                       y = value,
                       fill = variable)) +
    geom_bar(stat = 'identity') +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = value, label = value),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("Where do you obtain knowledge regarding NF1 should you have questions?",
            "Faceted by mode of inheritance") +
    ylab("Count")
#dev.off()
```

See the file ObtainKnowledgeByModeOfInheritance.png, in the images directory.

\hfill

```{r explained_medical_aspects}
display.df <- grand.data %>%
  group_by(mode_of_inheritance) %>%
  summarize(
    Doctor = sum(did_a_doctor_explain_the_medial_aspects_of_nf1_to_you.),
    GeneticCounselor = sum(did_a_genetic_counselor_explain_the_medial_aspects_of_nf1_to_you.),
    FamilyMembers = sum(did_a_family_member_explain_the_medial_aspects_of_nf1_to_you.),
    NoOne = sum(did_no_one_explain_the_medial_aspects_of_nf1_to_you.),
    NotSure = sum(are_you_not_sure_who_explained_the_medial_aspects_of_nf1_to_you.))
display.df <- melt(display.df, 'mode_of_inheritance')

levels(display.df$variable) <- c('Doctor','Genetic Counselor','Family Members','No One','Not Sure')

#tiff(filename = '../images/MedicalExplanationByModeOfInheritance.png',
#     units = 'in',
#     width = 8,
#     height = 5,
#     res = 300)
ggplot(display.df, aes(x = reorder(variable, -value),
                       y = value,
                       fill = variable)) +
    geom_bar(stat = 'identity') +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = value, label = value),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("Who explained the medical aspects of NF1 to you?",
            "Faceted by mode of inheritance") +
    ylab("Count")
#dev.off()
```

See the file MedicalExplanationByModeOfInheritance.png, in the images directory.

\hfill

```{r pairwise_age}
display.df <-  grand.data[grand.data$at_what_age_were_you_diagnosed_with_nf1. != "", c('at_what_age_were_you_diagnosed_with_nf1.','age_of_diagnosis')]

colnames(display.df) <- c('Parent','Child')
display.df <- display.df[display.df$Parent != 'Unknown',]
display.df$Parent <- as.numeric(display.df$Parent)

#tiff(filename = '../images/AgeOfDiagnosisFamilialRelationship.png',
#    units = 'in',
#    width = 8,
#     height = 5,
#     res = 300)
ggpaired(display.df,
         cond1 = 'Parent',
         cond2 = 'Child',
         fill = 'condition',
         palette = 'jco',
         xlab = "",
         ylab = "Age of Diagnosis",
         ggtheme = theme_gray()) +
  theme(legend.position = 'none') +
  ggtitle("Familial Relationship: Age of Diagnosis")
#dev.off()
```

See the file AgeOfDiagnosisFamilialRelationship.png, in the images directory.

# References

[1] McDonald, J.H. 2014. Handbook of Biological Statistics (3rd ed.). Sparky House Publishing, Baltimore, Maryland.