---
title: "NF1 Thesis Analysis"
author: "Marshall McQuillen"
date: "12/10/2018"
output: 
  pdf_document: 
    latex_engine: xelatex
    toc: true
    toc_depth: 4
---

```{r, include=FALSE, eval=TRUE}
# set global options
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE)
```

# Study Aims

1. To assess whether parental knowledge of NF1 is dependent on the child's mode of inheritance of the condition (familial vs de novo mutations)

2. To determine how exposure to genetic counseling affects knowledge of NF1.

```{r test_setup}
# load packages
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ltm))

# load & setup data
grand.data <- read.csv("/Users/marshallmcquillen/Data_Science/NF1_Thesis/data/grand_data.csv")
grand.data$mode_of_inheritance <- ifelse(grand.data$does_your_childs_other_parent_have_nf1. == "No" & grand.data$do_you_have_nf1. == "No", "De Novo", "Inherited")

# original significance level
alpha <- 0.05

# Bonferroni Correction
adj.alpha <- alpha / 4

# create subsets on whether respondent has seend a genetic counselor
grand.data$seen_gc <- grand.data$have_you_ever_met_with_a_genetic_counselor.
seen.gc <- subset(grand.data,
                  seen_gc == "Yes")
not.seen.gc <- subset(grand.data,
                      seen_gc == "No")

# create subsets based on whether the respondent has NF1
group.a <- subset(grand.data, group_id == "Group A")
group.b <- subset(grand.data, group_id == "Group B")
group.c <- subset(grand.data, group_id == "Group C")
```

## Data Overview & Cleaning

There were a total of 303 responses to the questionare:

* 11 respondents were "Not sure" if they had NF1. These respondents were dropped from the data set since both study aims are contingent upon knowing this information.

* One respondent skipped the question, "Does your child's other parent have NF1?" This respondent was kept since they answered the question, "Do you have NF1?" and their answer to "Does your child's other parent have NF1?" was imputed as "No."

* One respondent skipped the question, "Doe you have NF1?" This respondent was dropped from the data set since both study aims are contingent upon knowing this information.

* 14 respondents do not have NF1 and are "Not sure," if the childs other parent has NF1. These respondents were dropped from the data set since both aims of the thesis are dependent on this information.

* Three additional respondents were dropped from the study due to leaving other questions that will be discussed blank.

```{r group_summary_setup}
summary.df <- data.frame(Mean = c(mean(group.a$score),
                                  mean(group.b$score),
                                  mean(group.c$score)),
                         StandardDeviation = c(sd(group.a$score),
                                               sd(group.b$score),
                                               sd(group.c$score)))
rownames(summary.df) <- c('Group A','Group B','Group C')

age.range <- range(grand.data$age, na.rm = TRUE)
age.mean <- mean(grand.data$age, na.rm = TRUE)
age.sd <- sd(grand.data$age, na.rm = TRUE)
```

\pagebreak

After the above adjustments, the final data set contains 274 respondents, with an age ranging from `r age.range[1]` to `r age.range[2]` and a mean age of `r age.mean`. The standard deviation of the ages is `r age.sd`.

## Testing Methodology

Although there are only two separate study aims, the first will require a total of 3 pairwise statistical tests, comparing the average test scores of the three groups below to one another using Welch's $t$-Test.

* **Group A** - Respondents who do not have NF1 and the child's other parent also does not have NF1.

* **Group B** - Respondents who do not have NF1 and the child's other parent does have NF1.

* **Group C** - Respondents who do have NF1 and the child's other parent does not have NF1.

These three tests, combined with the test for the second study aim, create a total of 4 statistical tests. In order to control the family wise error rate, the original significance level ($\alpha = 0.05$) was adjusted using the Bonferroni Correction, reducing the level of statistical significance to $0.05 / 4 = 0.0125$.

A table with summar statistcs of the three groups is displayed below.

```{r group_summary_table}
knitr::kable(summary.df,
             caption = "Test Score Statistics")
```

\pagebreak

### Study Aim 1 - Does the mode of inheritance (familial vs de novo mutations) of the child have an effect on parental knowledge of NF1?

#### Test 1

* $H_0:$ Respondents in Group B do not have a different average test score than respondents in Group A.

* $H_a:$ Respondents in Group B have a different average test score than respondents in Group A.

\hfill

```{r}
# welch's t-test
test.1.results <- t.test(group.a$score, group.b$score, var.equal = FALSE, alternative = 'two.sided')

test.1.data <- rbind.data.frame(group.a, group.b)

subtitle <- paste("Mean's of each group displayed as vertical lines: Group A =", round(mean(group.a$score), 4), "|", "Group B =", round(mean(group.b$score), 4))

ggplot(test.1.data, aes(score, fill = group_id)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(group.a$score), color = '#F8766D') +
    geom_vline(xintercept = mean(group.b$score), color = '#00BFC4') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    ggtitle("Group A vs. Group B",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")


print(paste("Test 1 p value =", test.1.results$p.value))
```

Welch's two-tailed $t$-Test returns the above p value, with which we fail to reject the null hypothesis.

\pagebreak

#### Test 2

* $H_0:$ Respondents in Group A do not have a higher average test score than respondents in Group C.

* $H_a:$ Respondents in Group A have a higher average test score than respondents in Group C.

\hfill

```{r}
# welch's t-test
test.2.results <- t.test(group.a$score, group.c$score, var.equal = FALSE, alternative = 'greater')

test.2.data <- rbind.data.frame(group.a, group.c)

subtitle <- paste("Mean's of each group displayed as vertical lines: Group A =", round(mean(group.a$score), 4), "|", "Group C =", round(mean(group.c$score), 4))

ggplot(test.2.data, aes(score, fill = group_id)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(group.a$score), color = '#F8766D') +
    geom_vline(xintercept = mean(group.c$score), color = '#00BFC4') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    ggtitle("Group A vs. Group C",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")

print(paste("Test 2 p value =", test.2.results$p.value))
```

Welch's one-sided $t$-Test returns the above p value. Being well below the Bonferroni corrected significance level of `r adj.alpha`, the null hypothesis is rejected in favor of the alternative.

\pagebreak

#### Test 3

* $H_0:$ Respondents in Group B do not have a higher average test score than respondents in Group C.

* $H_a:$ Respondents in Group B have a higher average test score than respondents in Group C.

\hfill

```{r}
# welch's t-test
test.3.results <- t.test(group.b$score, group.c$score, var.equal = FALSE, alternative = 'greater')

test.3.data <- rbind.data.frame(group.b, group.c)

subtitle <- paste("Mean's of each group displayed as vertical lines: Group B =", round(mean(group.b$score), 4), "|", "Group C =", round(mean(group.c$score), 4))

ggplot(test.3.data, aes(score, fill = group_id)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(group.b$score), color = '#F8766D') +
    geom_vline(xintercept = mean(group.c$score), color = '#00BFC4') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    ggtitle("Group B vs. Group C",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")

print(paste("Test 3 p value =", test.3.results$p.value))
```

Welch's one-sided $t$-Test returns the above p value. Being well below the Bonferroni corrected significance level of `r adj.alpha`, the null hypothesis is rejected in favor of the alternative.

\pagebreak

#### Test 4

* $H_0:$ Respondents who have seen a genetic counselor do not have a higher average test score than those that have not seen a genetic counselor.

* $H_a:$ Respondents who have seen a genetic counselor have a higher average test score than those that have not seen a genetic counselor

\hfill

```{r}
# welch's t-test
test.4.results <- t.test(seen.gc$score, not.seen.gc$score, var.equal = FALSE,
       alternative = 'greater')

test.4.data <- rbind.data.frame(seen.gc, not.seen.gc)

subtitle <- paste("Mean's of each group displayed as vertical lines: Yes =", round(mean(seen.gc$score), 4), "|", "No =", round(mean(not.seen.gc$score), 4))

ggplot(test.4.data, aes(score, fill = seen_gc)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = mean(seen.gc$score), color = '#F8766D') +
    geom_vline(xintercept = mean(not.seen.gc$score), color = '#00BFC4') +
    theme(legend.position="bottom") +
    theme(legend.title=element_blank()) +
    ggtitle("Have you seen a Genetic Counselor?",
            subtitle = subtitle) +
    xlab("Test Score") +
    ylab("Density")

print(paste("Test 4 p value =", test.3.results$p.value))
```

Welch's one-sided $t$-Test returns the above p value. Being well below the Bonferroni corrected significance level of `r adj.alpha`, the null hypothesis is rejected in favor of the alternative.


```{r cronbachs_alpha, include=FALSE}
quiz.questions <- c('one_half_or_50._of_genetic_information_is_passed_down_from_mother_to_child',
                    'there_is_more_than_one_gene_that_causes_nf1',
                    'a_father_can_pass_down_an_nf1_gene_mutation_to_his_daughters',
                    'all_people_who_have_an_nf1_gene_mutation_will_develop_cancer',
                    'nf1_symptoms_vary_from_one_person_to_another',
                    'a_person_is_born_with_nf1',
                    'nf1_can_skip_generations',
                    'if_a_woman_with_nf1_has_one_child_with_nf1_her_second_child_will_definitely_not_have_nf1',
                    'people_with_nf1_are_generally_more_likely_to_develop_cancer_at_a_younger_age',
                    'if_a_woman_with_nf1_has_scoliosis_then_her_child_will_also_develop_scoliosis',
                    'tumors_in_nf1_can_appear_anywhere_in_the_body',
                    'people_with_nf1_from_different_families_will_always_have_different_symptoms',
                    'cafe.au.lait_spots_.brown_marks_on_the_skin._are_often_the_first_sign_that_a_person_has_nf1',
                    'a_baby_with_nf1_may_be_born_with_a_tumor',
                    'all_cases_of_nf1_can_be_detected_in_a_womans_pregnancy_by_ultrasound',
                    'half_of_people_with_nf1_have_a_family_history_of_nf1',
                    'a_person_with_nf1_can_develop_tumors_that_may_lead_to_vision_loss_or_blindness',
                    'women_with_nf1_are_at_an_increased_risk_for_breast_cancer',
                    'is_scoliosis_associated_with_nf1.',
                    'is_attention_deficit_hyperactivity_disorder_associated_with_nf1.',
                    'are_clubbed_feet_associated_with_nf1.',
                    'are_congenital_heart_defects_associated_with_nf1.',
                    'are_seizures_associated_with_nf1.',
                    'are_bumps_on_the_skin_associated_with_nf1.',
                    'are_allergies_associated_with_nf1.',
                    'is_high_blood_pressure_associated_with_nf1.',
                    'are_learning_disabilities_associated_with_nf1.',
                    'are_optic_gliomas_associated_with_nf1.',
                    'is_infertility_associated_with_nf1.',
                    'are_lisch_nodules_.dark_spots_on_the_iris._the_colored_part_of_the_eye._associated_with_nf1.',
                    'is_small_head_size_associated_with_nf1.',
                    'are_cataracts_associated_with_nf1.',
                    'how_is_nf1_diagnosed.',
                    'what_types_of_exams_may_be_helpful_when_trying_to_make_a_diagnosis_of_nf1.',
                    'what_causes_nf1.',
                    'what_is_the_function_of_the_nf1_gene_in_the_body.',
                    'are_men_or_women_more_likely_to_be_affected_by_nf1.',
                    'how_many_copies_of_a_non.working_nf1_gene_must_one_have_to_be_affected_with_nf1.',
                    'if_a_person_with_nf1_has_a_child_what_is_the_chance_that_the_child_will_have_nf1.',
                    'what_is_the_increased_risk_for_a_person_with_nf1_to_develop_cancer_in_their_lifetime_compared_to_someone_without_nf1.',
                    'during_childhood_what_kind_of_screening_may_be_done_for_optic_gliomas_.eye_tumors..',
                    'how_do_the_symptoms_of_nf1_change_as_a_person_gets_older.',
                    'what_is_the_most_concerning_symptom_of_a_tumor_to_suggest_it_has_become_malignant_.cancerous..')

quiz.df <- grand.data[, quiz.questions]
cronbachs <- cronbach.alpha(quiz.df)$alpha
```

### Cronbach's Alpha

Using 43 quiz questions returned a Cronbach Alpha value of `r cronbachs`.


### Test Limitations

**EXPLANATION OF TEST LIMITATION (MARSHALL NEEDS TO FINISH)**

```{r}
shapiro.test(group.a$score)
shapiro.test(group.b$score)
shapiro.test(group.c$score)
shapiro.test(seen.gc$score)
shapiro.test(not.seen.gc$score)
```

\pagebreak

### Demographic Analysis

```{r proportion_of_respondents_who_have_nf1}
ggplot(grand.data, aes(do_you_have_nf1., fill = do_you_have_nf1.)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Do you have have NF1?") +
    ylab("Proportion")

result <- table(grand.data$do_you_have_nf1.)
knitr::kable(result,
             caption = "Do you have NF1?",
             col.names = c('Response','Count'))
```

\hfill

```{r proportion_of_respondents_who_have_met_a_gc}
ggplot(grand.data, aes(have_you_ever_met_with_a_genetic_counselor.,
                       fill = have_you_ever_met_with_a_genetic_counselor.)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Have you ever met with a Genetic Counselor?") +
    ylab("Proportion")

result <- table(grand.data$have_you_ever_met_with_a_genetic_counselor.)
knitr::kable(result,
             caption = "Have you ever met with a Genetic Counselor?",
             col.names = c('Response','Count'))
```

\hfill

```{r employment_status}
grand.data$employment_status <- sub("^$", "Unknown", grand.data$employment_status)

ggplot(grand.data, aes(employment_status,
                       fill = employment_status)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Employment Status") +
    ylab("Proportion")

result <- table(grand.data$employment_status)
knitr::kable(result,
             caption = "What is your employment status?",
             col.names = c('Response','Count'))
```

\hfill

```{r gender}
grand.data$gender <- sub("^$", "Unknown", grand.data$gender)

ggplot(grand.data, aes(gender,
                       fill = gender)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Gender") +
    ylab("Proportion")

result <- table(grand.data$gender)
knitr::kable(result,
             caption = "Gender",
             col.names = c('Response','Count'))
```

\hfill

```{r ethnicity}
grand.data$ethnicity <- sub("^$", "Unkown", grand.data$ethnicity)

ggplot(grand.data, aes(ethnicity,
                       fill = ethnicity)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Ethnicity") +
    ylab("Proportion")

result <- table(grand.data$ethnicity)
knitr::kable(result,
             caption = "Ethnicity",
             col.names = c('Response \t','Count'))
```

\pagebreak

One person did not answer the question regarding their educational background, and therefore was excluded from the below plot.

```{r educational_background}
display.df <- grand.data[grand.data$educational_background != "", ]

ggplot(display.df, aes(educational_background,
                       fill = educational_background)) +
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    geom_text(aes(y = ((..count..)/sum(..count..)),
                  label = scales::percent((..count..)/sum(..count..))),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Educational Background") +
    ylab("Proportion")

result <- table(display.df$educational_background)
knitr::kable(result[2:length(result)],
             caption = "Educational Background",
             col.names = c('Response \t','Count'))
```

\pagebreak

It is important to note that the answers to the below question, "What specialists does your affected child see?" are not mutually exclusive - children who see a Neurologist may have seen an Oncologist and a Geneticist as well. 

\hfill

```{r specialists_seen}
cols = c("does_your_oldest_child_see_a_neurologist_for_nf_care.",
         "does_your_oldest_child_see_an_oncologist_for_nf_care.",
         "does_your_oldest_child_see_a_geneticist_for_nf_care.",
         "does_your_oldest_child_see_an_ent.audiologist_for_nf_care.",
         "does_your_oldest_child_see_a_pediatrician_for_nf_care.",
         "does_your_oldest_child_see_a_dermatologist_for_nf_care.",
         "mode_of_inheritance")

display.df <- grand.data %>%
    group_by(mode_of_inheritance) %>%
    summarise(Neurologist = sum(does_your_oldest_child_see_a_neurologist_for_nf_care.),
              Oncologist = sum(does_your_oldest_child_see_an_oncologist_for_nf_care.),
              Geneticist = sum(does_your_oldest_child_see_a_geneticist_for_nf_care.),
              Audiologist = sum(does_your_oldest_child_see_an_ent.audiologist_for_nf_care.),
              Pediatrician = sum(does_your_oldest_child_see_a_pediatrician_for_nf_care.),
              Dermatologist = sum(does_your_oldest_child_see_a_pediatrician_for_nf_care.))
df.d <- melt(display.df, id.vars='mode_of_inheritance')


ggplot(df.d, aes(x = variable, y = value, fill = variable)) +
    geom_bar(stat = 'identity') +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = value, label = value),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("What specialists does your affected child see?",
            "Faceted by mode of inheritance") +
    ylab("Count")
```

\pagebreak

```{r frequency_of_visits}
ggplot(grand.data, aes(how_often_does_your_.oldest._child_see_an_nf_doctor.,
                       y = (..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))) +
    geom_bar(aes(fill = how_often_does_your_.oldest._child_see_an_nf_doctor.)) +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(label = round(((..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))*100,2)),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("How often does your oldest child see a NF1 doctor?") +
    ylab("Proportion")

result <- table(grand.data$how_often_does_your_.oldest._child_see_an_nf_doctor., grand.data$mode_of_inheritance)
knitr::kable(result,
             caption = "How often does your oldest child see a NF1 doctor?")
```

\hfill

```{r frequency_of_visits_vs_score}
display.df <- grand.data %>%
    group_by(how_often_does_your_.oldest._child_see_an_nf_doctor.) %>%
    summarise(mean_score = mean(score))
df.d <- melt(display.df, id.vars='how_often_does_your_.oldest._child_see_an_nf_doctor.')


ggplot(df.d, aes(x = how_often_does_your_.oldest._child_see_an_nf_doctor., y = value, fill = how_often_does_your_.oldest._child_see_an_nf_doctor.)) +
    geom_bar(stat = 'identity') +
    scale_y_continuous(labels = scales::percent) +
    geom_text(aes(y = value, label = round(value*100, 2)),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("How often does your oldest child see a NF1 doctor?") +
    ylab("Mean Quiz Score")
```

\hfill

```{r mode_of_inheritance_vs_seen_gc}
ggplot(grand.data, aes(have_you_ever_met_with_a_genetic_counselor.,
                       fill = have_you_ever_met_with_a_genetic_counselor.)) +
    geom_bar(aes(y = (..count..))) +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = (..count..),
                  label = (..count..)),
              stat = "count",
              vjust = -0.25) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("Have you ever met with a Genetic Counselor?",
            "Faceted by mode of inheritance") +
    ylab("Count")
```

\hfill

```{r severity_symptoms}
ggplot(grand.data,
       aes(how_would_you_describe_the_severity_of_symptoms_of_your_.oldest._child_with_nf1.,
           y = (..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))) +
    geom_bar(aes(fill = how_would_you_describe_the_severity_of_symptoms_of_your_.oldest._child_with_nf1.)) +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(label = round(((..count..)/sapply(PANEL, FUN=function (x) sum(count[PANEL == x])))*100,2)),
              stat = "count",
              vjust = -0.25) +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.title = element_blank(),
          legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ggtitle("How would you describe the severity of symptoms\nof your oldest child with NF1?",
            "Faceted by mode of inheritance") +
    ylab("Proportion")

result <- table(grand.data$how_would_you_describe_the_severity_of_symptoms_of_your_.oldest._child_with_nf1., grand.data$mode_of_inheritance)
knitr::kable(result,
             caption = "How would you describe the severity of symptoms of your oldest child with NF1?")
```

\pagebreak

Similar to one of the previous plots, it is important to note that the answers to the below question, "Where do you obtain knowledge regarding NF1 should you have questions?" are not mutually exclusive - parents who ask their doctor might also reference online sources.

\hfill

```{r obtain_knowledge}
display.df <- grand.data %>%
  group_by(mode_of_inheritance) %>%
  summarize(
    Doctor = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_your_doctor.),
    NF1Organization = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_an_nf_organization_website.),
    FamilyMembers = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_family_members_with_nf1.),
    OnlineSources = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_online_searches.),
    SocialMedia = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_social_media_sites_.such_as_facebook..),
    OtherFamilies = sum(if_you_have_questions_regarding_nf1_do_you_obtain_knowledge_from_other_families_you_know_that_have_nf1.),
    DontInvestigate = sum(if_you_have_questions_regarding_nf1_do_you_not_obtain_additional_information.))
display.df <- melt(display.df, 'mode_of_inheritance')

ggplot(display.df, aes(x = variable, y = value, fill = variable)) +
    geom_bar(stat = 'identity') +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = value, label = value),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("Where do you obtain knowledge regarding NF1 should you have questions?",
            "Faceted by mode of inheritance") +
    ylab("Count")
```

\pagebreak

```{r explained_medical_aspects}
display.df <- grand.data %>%
  group_by(mode_of_inheritance) %>%
  summarize(
    Doctor = sum(did_a_doctor_explain_the_medial_aspects_of_nf1_to_you.),
    GeneticCounselor = sum(did_a_genetic_counselor_explain_the_medial_aspects_of_nf1_to_you.),
    FamilyMembers = sum(did_a_family_member_explain_the_medial_aspects_of_nf1_to_you.),
    NoOne = sum(did_no_one_explain_the_medial_aspects_of_nf1_to_you.),
    NotSure = sum(are_you_not_sure_who_explained_the_medial_aspects_of_nf1_to_you.))
display.df <- melt(display.df, 'mode_of_inheritance')

ggplot(display.df, aes(x = variable, y = value, fill = variable)) +
    geom_bar(stat = 'identity') +
    facet_grid(~mode_of_inheritance) +
    geom_text(aes(y = value, label = value),
              vjust = -0.25,
              size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'none',
          axis.title.x = element_blank()) +
    ggtitle("Who explained the medical aspects of NF1 to you?",
            "Faceted by mode of inheritance") +
    ylab("Count")
```